---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pwrml

<!-- badges: start -->
<!-- badges: end -->

This package offers implementation of analytical and sampling-based power analyses for the Wald, LR, score, and gradient tests in the frameworks of linear hypotheses and marginal maximum likelihood estimation.

The vignette "demo" gives an introduction of basic features, "adding_hypotheses" contains a tutorial on setting up custom hypotheses. Further details can be found in our paper "Power analysis for the Wald, LR, score, and gradient tests in a marginal maximum likelihood framework: Applications in IRT".

## Installation

```{r}
# library(devtools)
# install_github('anonymized')
```

## To-Dos:

- speed up calculation of score statistic using internals from the mirt functions, e.g. attributes(fitted$res)$Internals
