---
title: "Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demo}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```



```{r message=FALSE}
library(mmlpwrpackage)
library(mirt)
library(dplyr)
```

```{r eval=FALSE, include=FALSE}

# Benchmark Feature: Estimating the Time needed for Analytical Approach


# Ich würde ganz am Anfang erklären, an welche User sich die Vignette richtet. Anfänger, oder brauchen die Leute Erfahrung in mirt…?
# Ich würde auch jeden Chunk erklären, was da jeweils passiert.
# Für den Aufbau empfehle ich, dass du zuerst eine Einleitung mit Überblick bringst, dann ein Tutorial, wo du die Anwendung vorzeigst, und am Ende mögliche Erweiterungen.
# Welche Möglichkeiten für jedes Argument?
# Bessere Bezeichnungen, z.b. nicht ncps.


```


# Basic Features

## Usage with observed data

We want to know the power and required sample size for a test of the Rasch vs 2PL model. We use the LSAT 6 Dataset from the mirt package.


As a first step, we load the dataset and fit a 2PL model. The 2PL parameters are then used as parameters for the alternative hypothesis in our hypothesis definition. We do not have to provide the parameters under the null hypothesis here, because they are implicitly defined as the maximum likelihood parameters when slopes are equal. These maximum likelihood restricted parameters are calculated by the setup_hypothesis function.


```{r}
dat <- expand.table(LSAT7)
mirtfit <- mirt(dat,1,verbose = FALSE)
pars = coef_short(mirtfit)
pars
```

```{r}
hyp <- setup_hypothesis(type = "1PLvs2PL", altpars = pars)
```


For this hypothesis, we can estimate the power and required sample size.
```{r}
ncps <- calculate_ncps(hyp=hyp)
power(hyp=hyp,ncp=ncps,alpha=.05,ssize=500)
ssize(hyp=hyp,ncp=ncps,alpha=.05,power=.80)
```


## Usage without data

To apply a power analysis without data, we first setup the parameters that take the place of the estimated parameters in the application with data

```{r}
altpars <- list(
        a = rlnorm(5,sdlog = .4),
        d = rnorm(5)
        )
altpars
```

We can estimate the power and sample size as above.
```{r}
hyp <- setup_hypothesis(type = "1PLvs2PL", altpars = altpars)

ncps <- calculate_ncps(hyp=hyp)
power(hyp=hyp,ncp=ncps,alpha=.05,ssize=500)
ssize(hyp=hyp,ncp=ncps,alpha=.05,power=.80)
```


## Larger numbers of items: Simulated Ncp

Applying on same data as above. 
```{r}

ncps <- calculate_ncps(hyp=hyp,simbased = TRUE,n.pers = 10^5)
power(hyp=hyp,ncp=ncps,alpha=.05,ssize=500)
ssize(hyp=hyp,ncp=ncps,alpha=.05,power=.80)

```

New data with 50 items:

```{r}
altpars <- list(
        a = rlnorm(50,sdlog = .1),
        d = rnorm(50)
        )

hyp <- setup_hypothesis(type = "1PLvs2PL", altpars = altpars)

ncps <- calculate_ncps(hyp=hyp,simbased = TRUE,n.pers = 10^3)
power(hyp=hyp,ncp=ncps,alpha=.05,ssize=500)
ssize(hyp=hyp,ncp=ncps,alpha=.05,power=.80)

```


# Implemented Hypotheses

The following hypothesis are currently implemented.

* Rasch against 2PL
* DIF in 2PL

The Rasch against 2PL hypothesis is presented above. We will provide some examples for the other hypotheses. To implement custom hypothesis, see the section for package extensions. 


## DIF in 2PL

```{r}
group1 = group2 <- list(
        a = rlnorm(5,sdlog = .2),
        d = rnorm(5)
        )

group2$a[1] = (group2$a[1])^2
group2$d[1] = group2$d[1] + .5

altpars <- list(group1,group2)

altpars
```


```{r}
hyp <- setup_hypothesis(type = "DIF2PL", altpars = altpars)
ncps <- calculate_ncps(hyp=hyp)
power(hyp=hyp,ncp=ncps,alpha=.05,ssize=500)
ssize(hyp=hyp,ncp=ncps,alpha=.05,power=.80)


# experimental stuff:
# ncps2 <- calculate_ncps(hyp=hyp,simbased=TRUE,n.pers = 100000)
# 
# power(hyp=hyp,ncp=ncps2,alpha=.05,ssize=500)
# ssize(hyp=hyp,ncp=ncps2,alpha=.05,power=.80)


```



# Additional Features


## Generate artificial datasets

The hypothesis object can also be used to generate data according to the parameters of the alternative hypothesis. Note that setup.data also allows for non-normal person distributions, e.g. uniform or skewed-normal distributions.
```{r}
altpars <- list(
        a = rlnorm(5,sdlog = .4),
        d = rnorm(5)
        )

hyp <- setup_hypothesis(type = "1PLvs2PL", altpars = altpars)

data <- setup.data(hyp=hyp,n=500)
```


## Performing Hypothesis Tests

```{r}
fitted <- mml.fit(data = data,hyp = hyp)
```

From the results, we can calculate the observed statistics and p-values.

```{r}
stats_obs <- c(wald_obs(fitted),lr_obs(fitted),score_obs(fitted))
pvals <- pchisq(stats_obs,df=nrow(hyp$resmod$Amat),ncp=0,lower.tail=FALSE)
```


## Estimating the time needed for the analytical approach

We can estimate the time needed for the 1PL vs 2PL hypothesis and 10 items in seconds.
```{r}
h_1PLvs2PL$calctime(10)
```

Analogous calculation for the DIF hypothesis:
```{r}
h_DIF2PL$calctime(10)
```


## Extending the package


Present other hypotheses, their inner workings and make an example for an extension.






